---
title: "Analysis Report 2: Your Title Here"
author: "Don Francisco"
date: "November 21, 2018"
output: github_document
bibliography: references.bib
csl: bioinformatics.csl
---

# Introduction


Lung cancer is the leading cause of death among men and women in the United States [@torre2016lung]. Worldwide, it is also the leading cause of death among men and the second leading cause of death among women. Lung cancer rates and trends around the world and in the United States vary among different factors such as sex, age, race, socioeconomic status, and geography [@devesa2005international]. In the United States specifically, mortality rates in relation to lung cancer are highest among areas in the mid-South, males, black people, and people of lower socioeconomic status. Like any cancer, this can be caused by a number of factors such as genetics, environmental quality, and smoking status. 

In many studies that deal with lung cancer in the population, the smoking status of patients is most commonly looked at. In the year of 1954, William Watson and Alexander Conte, in their scientific study of looking at smoking and lung cancer mentioned how many clinical reports that looked at the relationship between tobacco smoking and lung cancer have looked into the fact that it is rare to find a lung cancer in a man who was not previously or currently a heavy smoker [@watson1954smoking]. In a more recent study as well, this fact is *still* being looked at. In the year of 2000, about 50 years later, Doll also emphasized how cigarette smoking is a significant factor in the production of carcinoma in the lung, along with all of the other harmful effects that it causes [@doll2000smoking]. 

Even with all of these factors that contribute to our possibilites of getting lung cancer, there are still lifestyle changes that we can make to help reduce these chances. One of the most straightforward life changes that we can work on making, is what we choose to include in our diets. Observational studies looking at the relationship between diet and lung cancer suggest that an increase in vegetable and fruit intake can reduce the risk in men and women, whether they are smokers, previous smokers, or people who have never smoked before [@ziegler1996nutrition]. But if this option among others do not work and lung cancer still prevails, there are many forms of treatment that can work to reduce the consequences related to the disease. While some forms of treatment are straightforward, such as quitting smoking and receiving palliative care, there are also more complicated treatments, like taking complex medications and getting surgery [@temel2010early].

Lung cancer rates, causes, reduction recommandations, and treatment can be difficult to pinpoint and understand as there are many factors that come into play. In "RNA-seq analysis of lung adenocarcinomas reveals different gene expression profiles between smoking and nonsmokin patients," Li et al. focused on exploring the gene expression profile and identifying the similarities and differences in looking at genes in smokers and nonsmokers [@li2015rna]. In looking at RNA-seq data from paired normal and tumor tissues in both 34 nonsmoking and 34 smoking patients with lung adenocarcinoma, they applied a general linear model to identify genes that were differentially expressed in nonsmoker and smoker patients and genes that varied between the two groups. Through their research, they discovered that geen expression varied at a considerable amount between nonsmoker and smoker patients with lung adenocarcinoma. They also discovered that smoking patients had more complicated disease mechanisms and abnormality with the regualation of metabolic processes. Through transcriptome analysis, their study revealed that there were pathogenetic differences in smoking and nonsmoking patients that had lung adenocarcinoma. 

In analyzing the data from Li et al.'s study, I looked at the variability in the expression of different genes. Using the TMPRSS4, SPP1, and MMP1 genes, which are genes that have been shown to be related with poor prognosis, more aggressive phenotype, or early onset of lung cancer, I had the following main question: With the gene that has the highest expression value within the three, how are the expression values of this gene affected in different patients and how do they vary in terms of cancer vs. normal tissue, the different stages of cancer, and the smoking status of the patients? In the Li et al. study, RNA-seq data was gathered from 34 nonsmoking patients and 34 smoking patients. Using this data, I trimmed the reads appropriately, ran the Sailfish tool for transcript quantification from RNA-seq data, and I built a transcript gene table. Taking a look at the gathered data, I hypothesized that the gene with the highest expression value between the three genes would also have higher expression values in tumor tissue compared to normal tissue, higher stages of the lung cancer, and in patients that have or are currently smoking. Seeing this result could ultimately lead to information that can benefit everyone. Specifically, as lung cancer is the leading cause of cancer death in the United States, understanding how genes and their expression patterns work, could give the population a better idea of how to possibly control this in the future. Specifically, if the upregulation of a certain gene is seen in tumor tissues, higher cancer stages, and positive smoking status, then we may be able to understand that this gene may be one of the introductions to lung caner. Overall, the analysis behind these genes and the data found in the Li et al. study can give us some insight about how to get more information about lung cancer in general. 

# Methods

## Sample origin and sequencing

### RNA Sequencing Data and Phenotype Data

Li et al. downloaded the transcriptome sequencing data from 68 lung adenocarcinoma patients with validated smoking status from Gene Expression Omnibus (GEO) with the accession number GSE40419 [@li2015rna]. In the 68 patients, both the tumor and paired normal tissue had been sequenced. With 34 patients who had never been smokers and 34 patients who either have smoked or are current smokers, they restricted analysis of the people who either have smoked or are current smokers to patients under 75 in order to ensure comparability in age ranges between both groups. Li et al. downloaded independent RNA-seq data from six patients who have never smoked from GEO with the accession number GSE37765 in order to validate the genes in the nonsmoker group. The age range for these six patients varied from 44 to 70. 

### Differential Gene Expression Analysis 

Li et al. conducted geneotype calling of single nucleotide variants (SNVs) on 136 RNA-seq bam files from 68 individuals [@li2015rna]. This was done in order to confirm the paired relationship of RNA-seq data from both normal and tumor tissue. In a 2012 study, Seo et al. used a clustering method in order to recognize differentially expressed genes in both smoker and nonsmoker patients that had lung adenocarcinoma. They discovered genes with varied expression between normal and tumor tissue across their data samples [@seo2012transcriptional]. Li et al. on the other hand, analyzed differential gene expression using pair-wise comparison between both normal and tumor tissue and nonsmokers and smokers. They calculated pair-wise concordance rates for 9180 pairs total, which includes between individuals and between individual pairs. They found that the there is a paired nature of RNA-seq data from tumor and normal tissue. Using Tophat, Li et al. aligned pair-end RNA-seq reads to human genome assembly Ensembl GRCh37. They used R Bioconductor edgeR to perform the differential expression analysis and applied a general lienar model (lung tissue expression~smoking+smoking:patient+smoking:tissue). This model allowed them to identify genes that were differentially expressed in normal versus tumor tissue in nonsmoker or smoker patients, as well as genes that act differently between smoker and nonsmoker patients. They only kept tags that had at least 1 count per million in at least half of the sample size in the analysis. They noted significant genes as genes with Benjamini-Hochberg adjusted FDF < 0.05 and absolute values of logFC greater than 1. 

### Gene Function Annotation and Pathway Analysis

In order to look into the functional roles and relationships of the genes with different expression in the analysis, Li et al. analyzed the pathway. To analyze the pathway, they submitted the logFC value of significant genes to Ingenuity Pathway Analysis (IPA). 

## Computational

These are the methods that were used to process the sequencing data. Should probably be at least a half of a page. At a very minimum should include citations for biomartr, trimmomatic, and sailfish. Note that these three methods references don't count towards the five references you need to cite in the introduction.

### Install Tools 

Given the data from the Li et al. study, I first downloaded important toolkits from NCBI. Specifically, I downloaded the newest precompiled sra-toolkit binary, which I then untarred it, unzipped it, and added the toolkit programs to PATH. I downloaded aspera for fast download of files from NCBI, installed it, and ran the program manually to configure download directory. After excising the column of sample run ids, I downloaded the target sra files from the file lise using aspera ascp. 

### Download Genome and Convert SRA to Fastq

Using R script, I installed the biomartr package fro mBioconductor and then used it to download a reference genome, transcripts, and annotations for the human genome from NCBI RefSeq. The biomartr package allows for straightforward function for bulk retrieval of the needed genomic data [@drost2017biomartr]. After downloading the genome from the Li et al. study, I converted the paired-end sra files to fastq files. 

### Build Sailfish Index 

Sailfish is a tool that is used for transcript quanitification from RNA-seq data [@patro2014sailfish]. This tool requires a set of target transcripts to quantify. With this requirement fulfilled, sailfish runs in two phases; indexing and quantification. For this analysis though, I installed the Sailfish source with the requirement of boost libraries, and built a Sailfish index by running this script: 

sailfish index -t _ncbi_downloads/RNA/Homo_sapiens_rna_from_genomic_refseq.fna -o output/sailfish-index

### Trimmomatic 

After running fastqc on all fastq files and saving the output to the output director, I trimmed the paired end reads in parallel. With the trimmomatics tool and paired end reads, adapters will be removed, leading low quality or N based will be removed, trailing low quality or N based will be removed, reads with a 4-base wide sliding window will be scanned and cut when the average quality per base drops below 15, and reads below 36 bases long will be dropped [@bolger2014trimmomatic]. I ran the trimmoatics tool by running this script: 

for file in "$@"; do
	TrimmomaticPE -threads 8 -basein $file \
		-baseout data/trimmed/$(basename -s _1.fastq $file).trim.fastq \
		ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 \
		SLIDINGWINDOW:4:20 MINLEN:50 &
done

After running this script to trim paired end reads, I checked if the output file for given input already exists. If it didn't, I ran Trimmomatic. This must be done because only so many threads can use Java at once, and so there may be files that are not yet trimmed. Such script was used to do so: 

for file in "$@"
do
	SAMPLE=$(basename -s _1.fastq $file)
	if [ `ls data/trimmed | grep -c $SAMPLE` -eq 0 ]
	then
		TrimmomaticPE -threads 8 -basein $file \
		  -baseout data/trimmed/$(basename -s _1.fastq $file).trim.fastq \
		  ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 \
		  SLIDINGWINDOW:4:20 MINLEN:50 &
	fi
done

### Running Sailfish 

In order to count all of the reads from the trimmed reads after both paired reads made it through Trimmomatic QC, I ran Sailfish using this script: 

for sample in "$@"
do
	sampleID=$(basename -s .trim_1P.fastq $sample)
	echo "--------------------------------------------------"
	echo "Processing sample $sampleID"
	sailfish quant -i output/sailfish-index/ \
		-l IU \
		-1 data/trimmed/$sampleID.trim_1P.fastq \
		-2 data/trimmed/$sampleID.trim_2P.fastq \
		-o output/sailfish_quants/$sampleID \
		-p 88
done

### Parse, Build Transcript Gene Table, and Consolidate Counts

In order to construct a transcript to gene ID mapping table, I parsed with the following script:

cat _ncbi_downloads/annotation/Homo_sapiens_genomic_refseq.gff \
	| perl -ane '$F[8]=~/Genbank:(\w+).*?gene=(\w+)/; print "$1\t$2\n";' \
	| sort | uniq | tail -n +2 > NCBI_GenbankID_to_gene_name.tsv
	
I then searched for the Genbank ID in the line of the input and cut out the full transcript ID for that match and adds the gene name after it in order to make our lookup table. 

I read in all of the transcript counts made by Sailfish individually on each sample. I combined these counts at the transcript level to the gene level, normalized them in terms of length and read depth, and then built a table with genes in the rows and samples in the columns.  I then combined the two other metadata files that were gathered from SRA and the supplementary table from the original manuscript. 

# Results

In addition to a minimum of 4-5 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion.

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("magrittr")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")
```

```{r load-data, message = FALSE, echo = FALSE}
# load the dataset from a compressed binary file
# it gets loaded as an object called "joined_table"
# this has 6.2 million rows...so you will need to be thoughtful about
# how you analyze the data so that you don't overwhelm your laptop
load("output/final_compiled_counts/joined_count_data.RData")

# test that it loaded correctly before proceeding
stopifnot(exists("joined_table"))
```

```{r make-summary-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
top_15 <- joined_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
top_15 %>%
  kable()
```

**Table 1**: The most highly expressed genes in both genders included *SFTPB* and *EEF1A1*.

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
top_15 %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
    geom_col(position = "dodge")
```

**Figure 1**: Here we show an example figure caption.

```{r make-boxplot-of-highly-expressed-genes, echo = FALSE}
# here we just want to pull out the unique gene names and turn
# them into a vector so we can use it below to make a boxplot
# we use the pull() funtion to get this as a vector, just like
# we did when making histograms several weeks ago
top_genes <- top_15 %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

# now we need to filter from the full data set again, because
# we don't just want summary data, we want all the data in
# order to make boxplots
joined_table %>%
  filter(genename %in% top_genes) %>%
  ggplot(aes(x = genename,
             y = counts_lengthscaledtpm,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~cancer_stage) +
    xlab("Gene Name") +
    ylab("Scaled read counts per gene") +
    ggtitle("Read counts per gene by gender and cancer stage") +
    theme_bw() + # simplifies theme
    theme(axis.text.x = # rotates x axis labels vertically
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 2**: Here we show another example figure caption.
```{r bar-graph-of-three-genes}
# This code creates a graph to look at the 
# comparison of expression values in these 
# three genes. These three genes are looked at 
# because they are the three genes that showed 
# variance between smokers and nonsmokers in the
# Li et al. study. The highest expression value 
# will be looked at in more detail in terms of its
# expression values in cancer tissue vs. normal tissue
# specific stages, and nonsmokers vs smokers. 
joined_table %>%
  filter(genename == "SERPIND1" | 
           genename == "TNNT1" | 
           genename == "PCSK2") %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = genename,
             y = mean_count))
    geom_col(position = "dodge")
```
**Figure 3** 

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity)
joined_table %>%
  filter (genename == "") %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = normal_or_cancer)) +
    geom_col(position = "dodge")
```
**Figure 4** 

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity)
joined_table %>%
  filter (genename == "") %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = cancer_stage)) +
    geom_col(position = "dodge")
```
**Figure 5** 

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity)
joined_table %>%
  filter (genename == "") %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = smoking_status)) +
    geom_col(position = "dodge")
```
**Figure 6**
# Discussion

Add around 1-2 pages interpreting your results and considering future directions one might take in analyzing these data.

Look up what the genes encode for (what protein they made) and how this can lead to cancer

# Sources Cited
